---
title: "BXLCuttingGarden - Workshop: ERP-GAMM"
author: "Marine Thieux"
date: "2026-09-16"
output: html_document
---

This workshop provides an introduction to generalized additive mixed models (GAMMs) for ERP data. The goal is to move from classical time-point based ERP analyses toward a continuous model of the waveform, while accounting for within-subject variability and temporal autocorrelation.

Useful references:

-   <https://noamross.github.io/gams-in-r-course/>

-   <https://neuropsychology.github.io/NeuroKit/studies/erp_gam.html>

-   <https://m-clark.github.io/generalized-additive-models/>

-   <https://www.taylorfrancis.com/books/mono/10.1201/9781315370279/generalized-additive-models-simon-wood>

**Why Generalized Additive (Mixed) Models - GAMM?**

GAMMs extend generalized linear models by allowing nonlinear relationships between variables. In ERP research, this is useful because the waveform is rarely linear over time: the signal may rise, peak, and return to baseline in a way that cannot be captured well by a single slope or a few predefined contrasts.

Compared with classical ERP analyses, GAMMs can:

-   account for temporal autocorrelation between nearby time points;

-   model smooth temporal trajectories instead of only a handful of time windows;

-   incorporate subject-level variability through random smooths;

-   reduce the need for multiple comparison corrections by using penalized smooths.

GAMMs are especially helpful when we want to ask not only *"Is there a difference at 200 ms?"* but also *"How does the whole ERP trajectory differ between conditions?"*

Instead of testing one time point at a time, we model the entire waveform as a smooth function of time, while accounting for repeated measures and subject variability.
This is the core idea behind GAMM-based ERP analysis: it preserves the temporal structure of the signal, handles non-linearity, and respects the repeated-measures nature of the data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the packages needed for data handling, plotting, and GAMM fitting

library(tidyverse)  # data wrangling and plotting tools
library(patchwork)  # combine plots in one panel
library(ggeffects)  # compute model-based predicted values for plotting
library(emmeans)    # estimated marginal means and simple contrasts
library(mgcv)       # main package for GAM and GAMM fitting
library(itsadug)    # extra tools for visualizing and interpreting GAMMs
library(gratia)     # plotting smooth terms from GAMM objects
library(modelbased) # extract summaries of model predictions
```

```{r}
# Read the ERP dataset (preprocessed ERP)

data <- read.csv("/Users/marinethieux/Desktop/EEG_CLEAN/simulated_ERP_GAMM_workshop.csv") %>%
  mutate(
    subject = as.factor(subject),
    condition = as.factor(condition),
    trial = as.factor(trial),
    channel = as.factor(channel)
  ) %>%
  arrange(subject, condition, trial, channel, time_ms) %>%
  group_by(subject, condition, trial, channel) %>%
  mutate(
    # AR_start identifies the first observation in a new autocorrelation sequence.
    # This is useful when fitting models with an AR(1) correlation structure.
    AR_start = row_number() == 1
  ) %>%
  ungroup()
```

```{r}
# Quick data checks (data structure/ balance of conditions).

data %>% summarise(
  participants = n_distinct(subject),
  channels = n_distinct(channel),
  trials = n_distinct(trial),
  observations = n())

table(data$condition)
```

# Classical ERP

```{r}
# Plot ERP averages by condition and channel
data %>%
  group_by(condition, channel, time_ms) %>%
  summarise(ampl_avg = mean(amplitude_uV), .groups = "drop") %>%
  ggplot(aes(x = time_ms, y = ampl_avg, color = channel)) +
  geom_hline(yintercept = 0) +
  geom_line(size = 0.3) +
  geom_vline(xintercept = 0) +
  theme(legend.position = "none") +
  facet_grid(row = vars(condition))
```

This first descriptive plot raises several questions:

-   Is there a difference between conditions?

-   If yes, when does it occur in time?

-   Does the shape of the response differ across conditions?

-   Is the effect consistent across electrodes?

-   How much of the observed difference is due to between-subject variability?

### Grand average

```{r}
# Grand-average ERP across all channels and trials

data %>%
  group_by(condition, time_ms) %>%
  summarise(
    ERP = mean(amplitude_uV),
    SD = sd(amplitude_uV),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time_ms, y = ERP, colour = condition)) +
  geom_ribbon(aes(x = time_ms, ymin = ERP - SD, ymax = ERP + SD, fill = condition),
              alpha = 0.2, size = 0.1) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Face" = "#FF5722", "Landscape" = "#2196F3")) +
  scale_fill_manual(values = c("Face" = "#FF5722", "Landscape" = "#2196F3")) +
  labs(x = "Time (ms)", y = "Amplitude (µV)", colour = "Condition", fill = "Condition") +
  theme_classic() +
  ggtitle("Grand average - all channels")
```

At this point, we might ask whether the amplitude differs between conditions at 100 ms, 230 ms, or another specific time point. But this is still a classical approach: it averages over trials, channels, and participants, and it relies on predefined timepoints.

```{r}
# Visualize the ERP by condition and electrode

data %>%
  group_by(condition, channel, time_ms) %>%
  summarise(
    amplitude = mean(amplitude_uV),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = time_ms, y = amplitude, colour = condition)) +
  scale_color_manual(values = c("Face" = "#FF5722", "Landscape" = "#2196F3")) +
  geom_line() +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ channel) +
  theme_classic()
```

The next question is: can we model the complete waveform as a smooth function of time, instead of testing a small set of selected time windows? To do that, we cannot use a simple linear effect. A spline introduces enough flexibility to represent a curved trajectory and estimate its shape from the data.

## Spline regression

```{r}
# Fit a spline regression (focusing on one electrode for demonstration). 

data_O1 <- data %>%
  filter(channel == "O1") %>%
  droplevels()

# The goal is to estimate a nonlinear temporal trajectory with a flexible basis.
# A spline basis allows the fitted curve to bend without requiring us to define fixed peaks or windows in advance.

model_spline <- lm(
  amplitude_uV ~ condition * splines::bs(time_ms,
                                         df = 10), # df controls the flexibility / bending of the spline
  data = data_O1)

summary(model_spline)
```

```{r}
# Create predictions from the spline model across a grid of time points

# We create a regular sequence of time values spanning the observed period.
# For each time point, the model predicts the expected ERP amplitude for each condition. 
# This allows a smooth trajectory to be displayed.

spline_grid <- expand.grid(
  time_ms = seq(min(data$time_ms), max(data$time_ms), length.out = 300),
  condition = levels(data$condition)
)

spline_grid$prediction <- predict(model_spline, newdata = spline_grid)

# The model gives a smooth estimated waveform rather than testing isolated timepoints one by one.
```

```{r}
ggplot(spline_grid, aes(x = time_ms, y = prediction, colour = condition)) +
  geom_line(linewidth = 1.1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Spline Regression",
    subtitle = "Nonlinear temporal ERP trajectories",
    x = "Time (ms)",
    y = "Predicted amplitude (µV)",
    colour = "Condition"
  ) +
  theme_classic()
```

We have moved from asking whether conditions differ at predefined time points to asking whether their ERP trajectories differ continuously over time. A GAM extends this idea by estimating smooth functions directly from the data, instead of forcing an *a priori* shape.

## GAMM

The GAMM is fitted to EEG data from -200 to 800 ms around stimulus onset, with ERP amplitude in O1 as the outcome variable. The main predictor is condition.

The term `s(time_ms, by = condition)` models the temporal trajectory separately for each condition. This is the key idea: each condition can have its own smooth shape over time. A penalty is then applied to control the wiggliness of the smooth, thereby avoiding overfitting while still allowing flexible curves.

```{r}
# First GAM: time-varying effect of condition without random smooths

# The `bam()` function is recommended for larger datasets because it is more efficient than `gam()`. Here, the model estimates:
#   - an overall mean difference between conditions;
#   - a smooth temporal effect that can differ by condition.

model <- mgcv::bam(
  amplitude_uV ~
    condition +
    s(time_ms, by = condition,
      k = 20,
      bs = "cr"),
  data = data_O1,
  method = "fREML"
)

# The `appraise()` and `draw()` functions help us check the fitted model visually.
appraise(model)

draw(model, residuals = FALSE) &
  theme_bw()
```

```{r}
# Plot predictions

# The parametric part of the model is the fixed effect of condition.
# The smooth part represents the nonlinear mapping between time and amplitude.

plot_parametric(model, pred = list(condition = c("Face", "Landscape")))
plot(ggemmeans(model, terms = c("time_ms", "condition")))
```

Individual deviations from the group-level pattern are introduced using a random smooth of time by subject.

This is useful because participants often differ not only in overall amplitude but also in the timing and shape of their ERP responses. Random smooths allow each subject to have their own trajectory while still estimating the shared group effect.

To handle possible outliers and non-normality, the model can use a scaled t distribution. Temporal autocorrelation can also be modeled with an AR(1) structure, which is important because adjacent time points in an ERP waveform are not independent.

```{r}
# GAMM with subject-level random smooths and AR(1) structure

# `s(time_ms, subject, k = 20, bs = "fs")` creates a factor-smooth interaction:
# each subject gets their own smooth trajectory over time.
# This captures subject-specific waveform shapes.
#
# `rho = 0.92` specifies an autocorrelation parameter for residuals.
# `AR.start` indicates when a new subject × condition × trial × channel series
# begins, so the autocorrelation is reset appropriately.

model.gamm <- mgcv::bam(
  amplitude_uV ~
    condition +
    s(time_ms, by = condition, k = 20, bs = "cr", m = 1) +
    s(time_ms, subject, k = 20, bs = "fs", m = 1),
  data = data_O1,
  rho = 0.92,
  AR.start = data_O1$AR_start,
  # family = "scat",  # optionally use a scaled-t distribution for robustness
  method = "fREML",
  nthreads = 12)
```

This model asks: *does the ERP trajectory differ between conditions while accounting for each participant’s own time-varying response profile?*

```{r}
# Examine the fitted GAMM

summary(model.gamm)
gam.check(model.gamm)

draw(model.gamm, residuals = FALSE) &
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
# Compare against a simpler nested model

# This model removes the condition-specific smooth, so the temporal pattern is
# assumed to be the same across conditions. By comparing models, we can test
# whether the condition-specific smooth adds explanatory value.

model.gamm.02 <- mgcv::bam(
  amplitude_uV ~
    condition +
    s(time_ms, k = 20, bs = "cr", m = 1) +
    s(time_ms, subject, k = 20, bs = "fs", m = 1),
  data = data_O1,
  rho = 0.92,
  AR.start = data_O1$AR_start,
  family = "scat",
  method = "fREML",
  nthreads = 12
)
```

We compare the two models (with and without the condition-specific smooth). Model selection is based on statistical comparison of fit and complexity, rather than only on raw significance at chosen timepoints.

```{r}
#  Model comparison
# The `compare_performance()` function gives a quick summary of model fit
# The `compareML()` provides a formal comparison of the nested models.

performance::compare_performance(model.gamm, model.gamm.02)

compareML(
  model.gamm,
  model.gamm.02,
  suggest.report = TRUE,
  print.output = TRUE,
  signif.stars = TRUE
)$table
```

```{r}
# Visualizing differences between conditions over time

# `plot_smooth()` displays model-adjusted trajectories for each condition.
# `plot_diff()` shows the estimated difference between conditions across time.
# When the simultaneous confidence interval excludes zero, the difference is
# considered meaningful over that temporal region.

par(mfrow = c(1, 2), cex = 1.1)

plot_smooth(
  model.gamm,
  view = "time_ms",
  rug = FALSE,
  plot_all = "condition",
  rm.ranef = FALSE,
  n.grid = 600,
  main = "with random")

plot_diff(
  model.gamm,
  view = "time_ms",
  comp = list(condition = c("Face", "Landscape")),
  rm.ranef = FALSE,
  n.grid = 600,
  sim.ci = TRUE)

par(mfrow = c(1, 2), cex = 1.1)

plot_smooth(
  model.gamm,
  view = "time_ms",
  rug = FALSE,
  plot_all = "condition",
  rm.ranef = TRUE,
  n.grid = 600,
  main = "without random")

plot_diff(
  model.gamm,
  view = "time_ms",
  comp = list(condition = c("Face", "Landscape")),
  rm.ranef = TRUE,
  n.grid = 600,
  sim.ci = TRUE)
```

We estimate the difference between conditions continuously across time. At each time point, the model yields an uncertainty interval around the difference. When the simultaneous 95% interval no longer overlaps zero, we interpret this as evidence for a difference in the ERP trajectory during that period. This is a more principled way to summarize continuous time-series effects than testing isolated time bins.

```{r}
# Pairwise contrasts at selected time points

# This bridges the GAMM approach to the more familiar classical ERP contrast interpretation: we can estimate condition differences at several times and inspect the contrasts.

emm <- emmeans(
  model.gamm,
  ~ condition | time_ms,
  at = list(time_ms = seq(from = 0.1, to = 0.8, by = 0.05))
)

pairs(emm) %>%
  as.data.frame() %>%
  knitr::kable()
```

